### 2018-02-23 submission/manuscript.Rmd
update overall results

@@

***Individual Taxa are Associated with Significant ORs for Carcinomas:*** 
Next we identified those taxa were associated with ORs that were significantly associated with having a normal colon or the presence of adenomas or carcinomas. No taxa had a significant OR for the presence of adenomas when we used data collected from stool or tissue samples (Table S3 & S4). In contrast, 8 taxa had significant ORs for the presence of carcinomas using data from stool samples. Of these, 4 are commonly associated with the oral cavity: Fusobacterium (OR = 0.XX (0.XX - 0.XX)), Parvimonas (OR = 0.XX (0.XX - 0.XX)), Porphyromonas (OR = 0.XX (0.XX - 0.XX)), and Peptostreptococcus (OR = 0.XX (0.XX - 0.XX)) [Table S3]. The other 4 were A  (OR = 0.XX (0.XX - 0.XX)), B (OR = 0.XX (0.XX - 0.XX)), C (OR = 0.XX (0.XX - 0.XX)), and D (OR = 0.XX (0.XX - 0.XX)). Among the data collected from tissue samples, only unmatched carcinoma samples had taxa with a significant OR. Those included Dorea (OR = 0.35 (0.22 - 0.55)), Blautia (OR = 0.47 (0.3 - 0.73)), and Weissella (OR = 5.15 (2.02 - 13.14)). Mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4]. For example, Fusobacterium had an OR of 3.98 (1.19-13.24); however, due to the small number of studies and considerable variation in the data, the Benjimani-Hochberg-corrected P-value was 0.93 [Table S4]. It is interesting to note that Ruminococcus and members of Clostridium group XI in stool and Dorea and Blautia in tissue had ORs that were significantly less than 1.0, which suggests that these populations are protective against the development of carcinomas. Overall, there was no overlap in the taxa with significant OR between stool and tissue samples.

***Mouth-Associated Genera are Associated with an Increased OR of Tumor:***
Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered OR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in OR using OTU richness and Shannon diversity. To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in OR. For both tissue and stool samples only ORs associated with an increase or decrease in carcioma tumors were significant after multiple comparison correction [Table S3 & S4]. Out of the `r length(crc_OR_stool$measure)` taxa that had significant ORs in stool samples `r length(filter(crc_OR_stool, measure %in% c("Fusobacterium", "Parvimonas", "Porphyromonas", "Peptostreptococcus"))[, "measure"])` were mouth-associated microbes. These mouth-associated genera significantly increased the ORs of carcinoma for stool samples and included *Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Table S3]. Conversely, mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4]. Only unmatched tissue samples had significant OR taxa and these were *Dorea* (OR = `r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "rr"], digits = 2)` (`r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_lb"], digits = 2)` - `r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_ub"], digits = 2)`)), *Blautia* (OR = `r round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "rr"], digits = 2)` (`r round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_lb"], digits = 2)` - `r round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_ub"], digits = 2)`)), and *Weissella* (OR = `r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "rr"], digits = 2)` (`r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_lb"], digits = 2)` - `r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_ub"], digits = 2)`)). Overall, there was little direct overlap in increased or decreased OR taxa between both tumor type and sample site.

***Select Community Models can Recapitulate Whole Community Models:***
Since specific genera increased the OR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera. We selected these genera based on significance after multiple comparison correction. If no taxa were significant after multiple comparison correction then no model for that specific grouping (i.e. adenoma stool) was analyzed. We first tested three model AUCs. These models were created using Random Forest and where either all genera, all OTUs, or significant OR taxa only. Next, the all genera models and any significant OR taxa models were tested across all studies that were not used to train the model. For stool, all models used had similar AUCs [Figure 2]. Although for adenoma and unmatched carcinoma this trend held, there were large differences in matched tissue based on whether all genera or OTUs were utilized [Figure S2]. When analyzing the tests sets that were comprised of genera data from other studies, both the all genera and singificant OR taxa only models had a similar ability to detect individuals with carcinomas, for both stool and tissue samples [Figure 3-S3].

In stool, the most common genera in the top 10 most important variables, in the full community models using all genera-based models, were *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 4]. Regardless of sample type, mouth-associated genera were present in models for carcinomas [Figure 4B & D]. Yet, none were present in the majority of studies and none were present in the adenoma models [Figure 4A & B]. For the full community OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 4C & 4D].

Unlike the stool-based Random Forest models, the tissue-based models, for the full genera from the first approach, showed no consistent representation of *Ruminococcaceae*, *Ruminococcus*, *Bacteroides*, and *Roseburia* in the top 10 most important model variables across study [Figure S4]. The vast majority of the top 10 model variables for the full community genera- and OTU-based models using colon tissue tended to be study specific. Further, there was very little overlap in the top 10 important variables between adenoma and carcinoma stage models, regardless of whether colon tissue or stool was used  [Figure S4]. This discordance between stool and colon tissue samples also applies to the mouth-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of matched versus unmatched colon tissue Random Forest models [Figure S4B-C & S4E-F]. This suggests that either the colon tissue microbiota is study or person dependent, that kit and/or other types of contamination associated with low biomass samples may be skewing the results, or that multiple microbes could act as the inflammatory stimulus needed to exacerbate mutations.

***CRC Studies are Underpowered for Detecting Small Effect Sizes:***
Next, we assessed how much confidence should be placed in the reported outcomes from each individual study by calculating the ability to detect a difference (power) and sample size needed for small, medium, and large effect size differences between cases and controls. When assessing the power of each study at different effect sizes the majority of studies achieved 80% power to detect a 30% or greater difference between groups [Figure 5A & B]. No study that we analyzed had the standard 80% power to detect an effect size difference equal to or below 10% [Figure 5A & B].  In order to achieve a power of 80%, for small effect sizes, studies used in our meta-analysis would need to recruit over 1000 individuals for both the case and control arms [Figure 5C]

>>>>

crc_matched_OR_tissue <- as.data.frame(read_csv("../data/process/tables/select_genus_OR_matched_tissue_composite.csv") %>% mutate(bh = p.adjust(pvalue, method = "BH")))

@@

***Individual Taxa are Associated with Significant ORs for Carcinomas:*** 
Next we identified those taxa were associated with ORs that were significantly associated with having a normal colon or the presence of adenomas or carcinomas. No taxa had a significant OR for the presence of adenomas when we used data collected from stool or tissue samples (Table S3 & S4). In contrast, `r length(crc_OR_stool$measure)` taxa had significant ORs for the presence of carcinomas using data from stool samples. Of these, 4 are commonly associated with the oral cavity: Fusobacterium (OR = `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`)), Parvimonas (OR = `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_ub"], digits = 2))`)), Porphyromonas (OR = `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_ub"], digits = 2))`)), and Peptostreptococcus (OR = `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_ub"], digits = 2))`)) [Table S3]. The other 4 were Clostridium XI (OR = `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_ub"], digits = 2))`)), Enterobacteriaceae  (OR = `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_ub"], digits = 2))`)), Escherichia (OR = `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_ub"], digits = 2))`)), and Ruminococcus (OR = `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_ub"], digits = 2))`)). Among the data collected from tissue samples, only unmatched carcinoma samples had taxa with a significant OR. Those included Dorea (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_ub"], digits = 2))`)), Blautia (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_ub"], digits = 2))`)), and Weissella (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_ub"], digits = 2))`)). Mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4]. For example, Fusobacterium had an OR of `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`; however, due to the small number of studies and considerable variation in the data, the Benjimani-Hochberg-corrected P-value was `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "bh"], digits = 2))` [Table S4]. It is interesting to note that Ruminococcus and members of Clostridium group XI in stool and Dorea and Blautia in tissue had ORs that were significantly less than 1.0, which suggests that these populations are protective against the development of carcinomas. Overall, there was no overlap in the taxa with significant OR between stool and tissue samples.

***Individual Significant OR Taxa Classify Carcinoma Poorly:***
Since specific taxa were associated with a significant OR for carcinoma we tested whether they would also be good classifiers of carcinoma. For stool, the `r length(crc_OR_stool$measure)` significant taxa did no better at classifying those with normal colons versus those with carcinomas than chance [Figure 3A]. Likewise in unmathced tissue samples the `r length(crc_unmatched_OR_tissue$measure)` significant taxa were no better than an AUC of 0.5 [Figure 3B]. These results suggest that although these taxa are significantly associated with a decreased or increased OR of carcinoma, individually they are poor classifiers of disease.

***Select Community Models can Recapitulate Whole Community Models:***
Since specific taxa increased or decreased the OR for carcinoma but performed poorly as an individual classifier of carcinoma and assessing ORs for adenoma we assessed whether the overall bacterial community was better at classifying disease. Three models were tested and included a full taxa model, full OTU model, and significant OR taxa model. If no taxa were significant after multiple comparison correction then no model for that specific grouping (i.e. adenoma stool) was analyzed. We first tested three model AUCs. Next, the all genera models and any significant OR taxa models were tested across all studies that were not used to train the model.

For stool, all models used had similar AUCs for both adenoma and carcinoma [Figure 4]. However, the adenoma AUCs were barely better than chance [Figure 4A]. Similarily, when analyzing the tests sets that were comprised of genera data from other studies the adenoma models performed poorly [Figure 5A]. The carcinoma models performed much better, with both the all genera and singificant OR taxa only models having a similar ability to detect individuals with carcinomas [Figure 5B]. The most common genera in the top 10 most important variables, in the full genera models used to classify carcinomas, were *Bacteroides* and *Lachnospiraceae* [Figure 6A]. Regardless of sample type, mouth-associated genera were present in models for carcinomas but not consistently across studies [Figure 6]. For the full community OTU-based models, both *Bacteroides*, *Blautia*, *Lachnospiraceae*, and *Ruminococcaceae* were present in the top 10 consistently across studies [Figure 6B]. Overall, these results suggest that multiple microbes could act as the inflammatory stimulus needed to exacerbate mutations leading each individual microbe as a poor individual classifier but much better in aggregate with others.

The tissue-based models had results that were dependent on genera or OTU level data and matched or unmatched samples [Figure S2]. Similarily to stool, the significant OR taxa performed as well as both the full taxa and OTU models [Figure S3B]. When analyzing the tests sets that were comprised of genera data from other studies, all models performed poorly with one exception being carcinoma classification in unmatched tissue [Figure S4]. Unlike stool, the test sets for the significant OR taxa did not perform as well as the all genera-based models [Figure S4C]. For the full genera models built using either matched or unmatched samples, showed no consistent taxa representation in the top 10 most important model variables across study [Figure S4A & S4B]. Conversely, the OTU-based models built using either matched or unmatched samples showed more commonalities [Figure S4C & S4D]. In models built using matched samples *Lachnospiraceae*, *Fusobacteriaceae*, *Comamonadaceae*, and *Bacteroidaceae* appeared in the top 10 percent in the majority of studies [Figure S4C]. Although models built with unmatched samples also had *Lachnospiraceae* and *Bacteroidaceae*, a major difference was the presence of *Ruminococcaceae* in the top 10 of every study [Figure S4D]. This results suggests that either the colon tissue microbiota is study or person dependent or that that kit and/or other types of contamination associated with low biomass samples may be skewing the results.

### 2018-02-23 submission/manuscript.tex
update overall results

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased OR
of Tumors:}} To assess differences in broad scale community metrics as
disease severity worsens Operational Taxonomic Unit (OTU) richness,
evenness, and Shannon diversity measurements were power transformed and
Z-score normalized. These metrics are commonly used to assess the total
number of OTUs, the equality of their abundance, and the overall
diversity, respectively. Using linear mixed-effect models to control for
study and variable region we assessed whether OTU richness, evenness, or
Shannon diversity changed in a step-wise manner with disease severity.
In stool, there was a significant decrease in both evenness and Shannon
diversity as disease severity moved from control to adenoma to carcinoma
(P-value = 0.025 and 0.043, respectively). We next tested whether the
detectable differences in community significantly increased in OR of
having an adenoma or carcinoma. For fecal samples, a decrease versus the
overall median in evenness resulted in a significantly increased OR for
carcinoma (OR = 1.66 (1.2 - 2.3), P-value = 0.0021) and adenoma (OR =
1.3 (1.02 - 1.65), P-value = 0.035) while a decrease versus the overall
median in Shannon diversity only increased the OR for carcinoma (OR =
1.61 (1.14 - 2.28), P-value = 0.0069) {[}Figure 1{]}. Using the
Bray-Curtis distance metric and PERMANOVA, it was also possible to
identify significant bacterial community changes, in specific studies,
for both carcinoma-associated and adenoma-associated microbiota versus
control {[}Table S1{]}.

Using similar transformations for tissue samples, linear mixed-effect
models were used on the transformed combined data to control for study,
re-sampling of the same individual, and 16S variable region to test
whether OTU richness, evenness, or Shannon diversity changed in a
step-wise manner as disease severity increased. For colon tissue, there
were no significant changes in OTU richness, evenness, or Shannon
diversity as disease severity progressed from control to adenoma to
carcinoma (P-value \textgreater{} 0.05). We next analyzed the OR, for
matched (unaffected tissue and an adenoma or carcinoma from the same
individual) and unmatched (control and adenoma or carcinoma tissue not
from the same individual) colon tissue samples. For individuals at
either an adenoma or carcinoma stage of disease there was no significant
change in OR based on lower than median values for OTU richness,
evenness, and Shannon diversity {[}Figure S1 \& Table S2{]}. Similar to
stool samples, significant differences in bacterial community, assessed
by PERMANOVA, were identified in unmatched tissue samples, for those at
either adenoma or carcinoma stage of CRC {[}Table S1{]}. For studies
with matched samples no differences in bacterial community were observed
when assessed with PERMANOVA {[}Table S1{]}. These tissue results
suggest that the microbiota within an individual are similar to each
other regardless of disease status.

\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
OR of Tumor:}} Next, we asked if being higher than the median relative
abundance, for any specific genera, resulted in an altered OR for
adenoma or carcinoma, in stool and colon tissue, due to our previous
observations of small increases in OR using OTU richness and Shannon
diversity. To investigate this we analyzed all common genera across each
study, in colon tissue or stool, and assessed whether a relative
abundance higher than the median results in an increase or decrease in
OR. For both tissue and stool samples only ORs associated with an
increase or decrease in carcioma tumors were significant after multiple
comparison correction {[}Table S3 \& S4{]}. Out of the 8 taxa that had
significant ORs in stool samples 4 were mouth-associated microbes. These
mouth-associated genera significantly increased the ORs of carcinoma for
stool samples and included \emph{Fusobacterium}, \emph{Parvimonas},
\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Table S3{]}.
Conversely, mouth-associated genera were not significantly associated
with an increased OR for carcinoma in tissue samples {[}Table S4{]}.
Only unmatched tissue samples had significant OR taxa and these were
\emph{Dorea} (OR = 0.35 (0.22 - 0.55)), \emph{Blautia} (OR = 0.47 (0.3 -
0.73)), and \emph{Weissella} (OR = 5.15 (2.02 - 13.14)). Overall, there
was little direct overlap in increased or decreased OR taxa between both
tumor type and sample site.

\textbf{\emph{Select Community Models can Recapitulate Whole Community
Models:}} Since specific genera increased the OR for carcinoma over
diversity metrics we assessed whether the bacterial community was better
at classifying disease versus only a select group of genera. We selected
these genera based on significance after multiple comparison correction.
If no taxa were significant after multiple comparison correction then no
model for that specific grouping (i.e.~adenoma stool) was analyzed. We
first tested three model AUCs. These models were created using Random
Forest and where either all genera, all OTUs, or significant OR taxa
only. Next, the all genera models and any significant OR taxa models
were tested across all studies that were not used to train the model.
For stool, all models used had similar AUCs {[}Figure 2{]}. Although for
adenoma and unmatched carcinoma this trend held, there were large
differences in matched tissue based on whether all genera or OTUs were
utilized {[}Figure S2{]}. When analyzing the tests sets that were
comprised of genera data from other studies, both the all genera and
singificant OR taxa only models had a similar ability to detect
individuals with carcinomas, for both stool and tissue samples {[}Figure
3-S3{]}.

In stool, the most common genera in the top 10 most important variables,
in the full community models using all genera-based models, were
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} {[}Figure
4{]}. Regardless of sample type, mouth-associated genera were present in
models for carcinomas {[}Figure 4B \& D{]}. Yet, none were present in
the majority of studies and none were present in the adenoma models
{[}Figure 4A \& B{]}. For the full community OTU-based models,
\emph{Ruminococcaceae} was present in the top 10 consistently for both
adenoma and carcinoma models while \emph{Roseburia} was only present in
many adenoma models and \emph{Bacteroides} was present in the
overwhelming majority of the carcinoma models {[}Figure 4C \& 4D{]}.

Unlike the stool-based Random Forest models, the tissue-based models,
for the full genera from the first approach, showed no consistent
representation of \emph{Ruminococcaceae}, \emph{Ruminococcus},
\emph{Bacteroides}, and \emph{Roseburia} in the top 10 most important
model variables across study {[}Figure S4{]}. The vast majority of the
top 10 model variables for the full community genera- and OTU-based
models using colon tissue tended to be study specific. Further, there
was very little overlap in the top 10 important variables between
adenoma and carcinoma stage models, regardless of whether colon tissue
or stool was used {[}Figure S4{]}. This discordance between stool and
colon tissue samples also applies to the mouth-associated genera with
one noticeable skew being that \emph{Fusobacterium} and
\emph{Fusobacteriaceae} occur more often in the top 10 of matched versus
unmatched colon tissue Random Forest models {[}Figure S4B-C \& S4E-F{]}.
This suggests that either the colon tissue microbiota is study or person
dependent, that kit and/or other types of contamination associated with
low biomass samples may be skewing the results, or that multiple
microbes could act as the inflammatory stimulus needed to exacerbate
mutations.

\textbf{\emph{CRC Studies are Underpowered for Detecting Small Effect
Sizes:}} Next, we assessed how much confidence should be placed in the
reported outcomes from each individual study by calculating the ability
to detect a difference (power) and sample size needed for small, medium,
and large effect size differences between cases and controls. When
assessing the power of each study at different effect sizes the majority
of studies achieved 80\% power to detect a 30\% or greater difference
between groups {[}Figure 5A \& B{]}. No study that we analyzed had the
standard 80\% power to detect an effect size difference equal to or
below 10\% {[}Figure 5A \& B{]}. In order to achieve a power of 80\%,
for small effect sizes, studies used in our meta-analysis would need to
recruit over 1000 individuals for both the case and control arms
{[}Figure 5C{]}

>>>>

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased
Odds Ratio (OR) of Tumors:}} We first assessed whether variation in
broad community metrics like total number of operational taxonomic units
(OTUs) (i.e.~richness), the evenness of their abundance, and the overall
diversity was associated with disease stage after controlling for study
and variable region differences. In stool, there was a significant
decrease in both evenness and diversity as disease severity progressed
from normal to adenoma to carcinoma (P-value = 0.025 and 0.043,
respectively) {[}Figure 1{]}; there was not a significant difference for
richness (P-value = 0.21). We next tested whether the decrease in these
community metrics translated into significant ORs for having an adenoma
or carcinoma. For fecal samples, the ORs for richness were not
significantly greater than 1.0 for adenoma or carcinoma (P-value = 0.40)
{[}Figure 2A{]}. The ORs for evenness were significantly higher than 1.0
for adenoma (OR = 1.3 (1.02 - 1.65), P-value = 0.035) and carcinoma (OR
= 1.66 (1.2 - 2.3), P-value = 0.0021) {[}Figure 2B{]}. The ORs for
diversity were only significantly greater than 1.0 for carcinoma (OR =
1.61 (1.14 - 2.28), P-value = 0.0069), but not for adenoma (P-value =
0.11) {[}Figure 2C{]}. Although these OR are significantly greater than
1.0, it is doubtful that these are clinically meaningful values.

Similar to our analysis of sequences obtained from stool samples, we
repeated the analysis using sequences obtained from colon tissue. There
were no significant changes in richness, evenness, or diversity as
disease severity progressed from control to adenoma to carcinoma
(P-value \textgreater{} 0.05). We next analyzed the OR, for matched
(i.e.~where unaffected tissue and tumors were obtained from the same
individual) and unmatched (i.e.~where unaffected tissue and tumor tissue
were not obtained from the same individual) tissue samples. The ORs for
adenoma and carcinoma by any measure were not significantly different
from 1.0 (P-value \textgreater{} 0.05) {[}Figure S1 \& Table S1{]}. This
is likely due to the combination of a small effect size, as suggested
from the results using stool, and the relatively small number of studies
and size of studies used in the analysis.

\textbf{\emph{Disease Progression is Associated with Community-Wide
Changes in Composition and Abundance:}} Based on the differences in
evenness and diversity, we next asked whether there were community-wide
differences in the structure of the communities associated with
different disease stages. We identified significant bacterial community
differences in the stool of patients with adenomas relative to those
with normal colons in 1 of 4 studies and in patients with carcinomas
relative to those with normal colons in 6 of 7 studies (PERMANOVA;
P-value \textless{} 0.05) {[}Table S2{]}. Similar to the analyses using
stool samples, there were significant differences in bacterial community
structure between subjects with normal colons and those with adenoma (1
of 2 studies) and carcinoma (1 of 3 studies) {[}Table S2{]}. For studies
that used matched samples no differences in bacterial community
structures were observed {[}Table S2{]}. Combined, these results
indicate that there consistent and significant community-wide changes in
the fecal community structure of subjects with carcinomas. However, the
signal observed in subjects with adenomas or when using tissue samples
was not as consistent. This is likely due to a smaller effect size or
the relatively small sample sizes among the studies that characterized
the tissue microbiota.

\textbf{\emph{Individual Taxa are Associated with Significant ORs for
Carcinomas:}} Next we identified those taxa were associated with ORs
that were significantly associated with having a normal colon or the
presence of adenomas or carcinomas. No taxa had a significant OR for the
presence of adenomas when we used data collected from stool or tissue
samples (Table S3 \& S4). In contrast, 8 taxa had significant ORs for
the presence of carcinomas using data from stool samples. Of these, 4
are commonly associated with the oral cavity: Fusobacterium (OR = 2.74
(1.95 - 3.85)), Parvimonas (OR = 3.07 (2.11 - 4.46)), Porphyromonas (OR
= 3.2 (2.26 - 4.54)), and Peptostreptococcus (OR = 7.11 (3.84 - 13.17))
{[}Table S3{]}. The other 4 were Clostridium XI (OR = 0.65 (0.49 -
0.86)), Enterobacteriaceae (OR = 1.79 (1.33 - 2.41)), Escherichia (OR =
2.15 (1.57 - 2.95)), and Ruminococcus (OR = 0.63 (0.48 - 0.83)). Among
the data collected from tissue samples, only unmatched carcinoma samples
had taxa with a significant OR. Those included Dorea (OR = 0.35 (0.22 -
0.55)), Blautia (OR = 0.47 (0.3 - 0.73)), and Weissella (OR = 5.15 (2.02
- 13.14)). Mouth-associated genera were not significantly associated
with an increased OR for carcinoma in tissue samples {[}Table S4{]}. For
example, Fusobacterium had an OR of 3.98 (1.19 - 13.24; however, due to
the small number of studies and considerable variation in the data, the
Benjimani-Hochberg-corrected P-value was 0.93 {[}Table S4{]}. It is
interesting to note that Ruminococcus and members of Clostridium group
XI in stool and Dorea and Blautia in tissue had ORs that were
significantly less than 1.0, which suggests that these populations are
protective against the development of carcinomas. Overall, there was no
overlap in the taxa with significant OR between stool and tissue
samples.

\textbf{\emph{Individual Significant OR Taxa Classify Carcinoma
Poorly:}} Since specific taxa were associated with a significant OR for
carcinoma we tested whether they would also be good classifiers of
carcinoma. For stool, the 8 significant taxa did no better at
classifying those with normal colons versus those with carcinomas than
chance {[}Figure 3A{]}. Likewise in unmathced tissue samples the 3
significant taxa were no better than an AUC of 0.5 {[}Figure 3B{]}.
These results suggest that although these taxa are significantly
associated with a decreased or increased OR of carcinoma, individually
they are poor classifiers of disease.

\textbf{\emph{Select Community Models can Recapitulate Whole Community
Models:}} Since specific taxa increased or decreased the OR for
carcinoma but performed poorly as an individual classifier of carcinoma
and assessing ORs for adenoma we assessed whether the overall bacterial
community was better at classifying disease. Three models were tested
and included a full taxa model, full OTU model, and significant OR taxa
model. If no taxa were significant after multiple comparison correction
then no model for that specific grouping (i.e.~adenoma stool) was
analyzed. We first tested three model AUCs. Next, the all genera models
and any significant OR taxa models were tested across all studies that
were not used to train the model.

For stool, all models used had similar AUCs for both adenoma and
carcinoma {[}Figure 4{]}. However, the adenoma AUCs were barely better
than chance {[}Figure 4A{]}. Similarily, when analyzing the tests sets
that were comprised of genera data from other studies the adenoma models
performed poorly {[}Figure 5A{]}. The carcinoma models performed much
better, with both the all genera and singificant OR taxa only models
having a similar ability to detect individuals with carcinomas {[}Figure
5B{]}. The most common genera in the top 10 most important variables, in
the full genera models used to classify carcinomas, were
\emph{Bacteroides} and \emph{Lachnospiraceae} {[}Figure 6A{]}.
Regardless of sample type, mouth-associated genera were present in
models for carcinomas but not consistently across studies {[}Figure
6{]}. For the full community OTU-based models, both \emph{Bacteroides},
\emph{Blautia}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} were
present in the top 10 consistently across studies {[}Figure 6B{]}.
Overall, these results suggest that multiple microbes could act as the
inflammatory stimulus needed to exacerbate mutations leading each
individual microbe as a poor individual classifier but much better in
aggregate with others.

The tissue-based models had results that were dependent on genera or OTU
level data and matched or unmatched samples {[}Figure S2{]}. Similarily
to stool, the significant OR taxa performed as well as both the full
taxa and OTU models {[}Figure S3B{]}. When analyzing the tests sets that
were comprised of genera data from other studies, all models performed
poorly with one exception being carcinoma classification in unmatched
tissue {[}Figure S4{]}. Unlike stool, the test sets for the significant
OR taxa did not perform as well as the all genera-based models {[}Figure
S4C{]}. For the full genera models built using either matched or
unmatched samples, showed no consistent taxa representation in the top
10 most important model variables across study {[}Figure S4A \& S4B{]}.
Conversely, the OTU-based models built using either matched or unmatched
samples showed more commonalities {[}Figure S4C \& S4D{]}. In models
built using matched samples \emph{Lachnospiraceae},
\emph{Fusobacteriaceae}, \emph{Comamonadaceae}, and
\emph{Bacteroidaceae} appeared in the top 10 percent in the majority of
studies {[}Figure S4C{]}. Although models built with unmatched samples
also had \emph{Lachnospiraceae} and \emph{Bacteroidaceae}, a major
difference was the presence of \emph{Ruminococcaceae} in the top 10 of
every study {[}Figure S4D{]}. This results suggests that either the
colon tissue microbiota is study or person dependent or that that kit
and/or other types of contamination associated with low biomass samples
may be skewing the results.
